---
title: "homework ii"
author: "Sachin Mohan Sujir"
date: '2020-09-11'
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
header-includes: \usepackage{booktabs}
---

```{r echo=FALSE}
# This chunk is just to make it possible to shrink the typeface in succeeding chunks. Mainly this will be used for the crosstabs.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Introduction
311 is a telephone number that is used for non-emergency government services. The dataset consists of more than 9 million records which has data of the service call requests reported in the New York city from the year 2010 to the present year.

# Initialization
Here we load the tidyverse packages and the `data.table` package and load the nyc311 data set. Then we fix the column names of the nyc311 data so that they have no spaces.

```{r initialize}
library(tidyverse)

library(data.table)
# fast for when you are starting out:
# nyc311<-fread("311_Service_Requests_from_2010_to_Present.csv",nrow=10000)
# after you get going:
nyc311<-fread("311_Service_Requests_from_2010_to_Present.csv")
names(nyc311)<-names(nyc311) %>%
  stringr::str_replace_all("\\s", ".")
```

# Data pre-processing
We perform data pre-processing by dropping irrelevant columns and also removing duplicate rows.

```{r dataPreProcessing}
names(nyc311)
nyc311 <- nyc311[,c(-1,-10:-19,-23, -25:-49)]
names(nyc311)
nyc311 <- distinct(nyc311)
dim(nyc311)
```

# Description
Here we describe the data, showing both a sample and a data dictionary.

## The head of the table
Here we produce a table of just some relevant columns of data.

```{r tabulate, results="asis"}

library(xtable)
options(xtable.comment=FALSE)
options(xtable.booktabs=TRUE)
narrow<-nyc311 %>%
  select(Agency,
	 Complaint.Type,
	 Descriptor,
	 Incident.Zip,
	 Status,
	 Borough)
xtable(head(narrow))
```


## Data Dictionary
•	Created.Date – The date when the service request was created. (Type: Date and Time)
•	Closed.Date – The date when the service request was closed by the responding agency. (Type: Data and Time)
•	Agency – Acronym of responding agency. (Plain Text)
•	Agency.Name – Full Agency name of responding City Government Agency. (Type: Plain Text)
•	Complaint.Type – The type of complaint reported (For example: vending, illegal parking, blocked driveway).
•	Descriptor - Detailed description of the corresponding complaint type. (Type: Plain Text)
•	Location.Type – The type of location based on the address information. (Plain Text)
•	Incident.Zip – Zip code of the incident location. (Type: Plain Text)
•	Status – The status of the service request submitted. (Type: Plain Text)
•	Due.Date – The date, during when the responding agency is expected to update the service request. (Type: Date and Time)
•	Resolution.Action.Updated.Date – Date when the responding agency last updated the service request. (Type: Date and Time)
•	Borough – town/ district of the NYC provided by submitter. (Values: BRONX, BROOKLYN, MANHATTAN, QUEENS, STATEN ISLAND) (Type: Plain Text)
•	Latitude – Geo-based latitude of the incident location. (Type: Number)
•	Longitude – Geo-based longitude of the incident location. (Type: Number)
•	Location – Combination of the geo-based latitude and longitude of the incident location. (Type: location)

```{r}
names(nyc311)
```

# Exploration
Here we explore the columns in the data set.

The following plot shows a horizontal bar chart showing the top agencies that received service call requests along with the count of service call requests for each agency. 
```{r explore}
bigAgency <- narrow %>%
  group_by(Agency) %>%
  summarize(count=n()) %>%
  filter(count>1000)
bigAgency$Agency<-factor(bigAgency$Agency,
  levels=bigAgency$Agency[order(bigAgency$count)])
p<-ggplot(bigAgency,aes(x=Agency,y=count)) +
   geom_bar(stat="identity") +
   coord_flip() +
  ggtitle("The number of Complaints received per Agency")
p
```
The following bar chart shows the top 10 complaint types received, with the color specified for each complaint type.

```{r}
options(scipen = 999)
topComplaints <- narrow %>%
  group_by(Complaint.Type) %>%
  summarize(count=n()) %>%
  filter(count>100000) %>%
  top_n(10)
topComplaints$Complaint.Type<-factor(topComplaints$Complaint.Type,
  levels=topComplaints$Complaint.Type[order(topComplaints$count)])
plotA<-ggplot(topComplaints,aes(x=Complaint.Type,y=count, fill=Complaint.Type)) +
   geom_bar(stat="identity") +
   coord_flip() + 
   ggtitle("Complaints per category")
plotA
```
The following shows a coxcomb plot showing the boroughs that received the most service call requests depicted in the form of coxComb.
```{r}
boroughs <- narrow %>%
  group_by(Borough) %>%
  summarize(count=n())
boroughs$Borough<-factor(boroughs$Borough,
  levels=boroughs$Borough[order(boroughs$count)])
plotB<-ggplot(boroughs,aes(x=Borough,y=count, fill=Borough)) +
   geom_bar(stat="identity", width=1) +
   theme(aspect.ratio =1) +
   coord_polar() +
   ggtitle("Complaints per borough")
plotB
```
Compaints That NYPD received across each borough
```{r}
hpdComplaints <-dplyr::filter(narrow, Agency=='NYPD') 
hpdComp <- hpdComplaints %>%
  group_by(Complaint.Type,Borough) %>%
  summarize(Complaints = length(Complaint.Type))
ggplot(hpdComp, aes(x=Complaint.Type,y=Complaints, fill=Borough)) + 
  xlab("Complaint Type") +
  geom_bar(stat ="identity") +
  coord_flip() +
  ggtitle("NYPD Complaints by category")
```

The table below shows information about the number of open and closed service call requests.

```{r statusPlot}
statusFrequency <- narrow %>%
  group_by(Status) %>%
  summarize(count=n()) %>%
  filter(Status=="Open" | Status=="Closed")
statusFrequency$Status<-factor(statusFrequency$Status,
  levels=statusFrequency$Status[order(statusFrequency$count)])
statusFrequency
```
The bar below shows the percentage complaint type that are open and closed service requests of top 5 complaints. It seems like "Plumbing" has too many open service requests abd "Heating" requets have a good record.

```{r complaintStatus}
filteredData <- dplyr::filter(narrow, (Complaint.Type=='HEATING' | Complaint.Type=='GENERAL CONSTRUCTION' | Complaint.Type=='PLUMBING' | Complaint.Type=='Street Condition' | Complaint.Type=='Street Light Condition') & (Status=='Open' | Status=='Closed'))
complaintStatus <- filteredData %>%
  group_by(Status,Complaint.Type) %>%
  summarize(count=n())
plotC<-ggplot(complaintStatus,aes(x=Status,y=count, fill=Complaint.Type)) +
   geom_bar(stat="identity", position = "fill")
plotC
```

Next we include a crosstabulation.

```{r crosstabs, size='footnotesize'}
xtabA<-dplyr::filter(narrow,
  Complaint.Type=='HEATING' |
  Complaint.Type=='GENERAL CONSTRUCTION' |
  Complaint.Type=='PLUMBING'
)
xtabB<-select(xtabA,Borough,"Complaint.Type")
library(gmodels)
CrossTable(xtabB$Borough,xtabB$'Complaint.Type')
```

The above crosstab shows tabulation of every borough with respect to the complaint types- heating, general construction and plumbing, that is it shows the number of complaints received in every borrough for the three specific complaint types and along with chi-square contribution, the percentage of complaints in every borough(N/row total), percentage of each complaint type(N/column total) and percentage of complaints for a specific complaint type and at a specific borough.(N/table total).


```{r crosstabsAgency, size='footnotesize'}
xtabA1<-dplyr::filter(narrow, ( Agency=='HPD' | Agency=='NYPD'))
xtabB1<-select(xtabA1,Borough, Agency)
library(gmodels)
CrossTable(xtabB1$Borough,xtabB1$Agency)
```
The above crosstab shows the number of service requests received by HPD and NYPD agencies with respect to each borough.
```{r geoPlot}
locationData <-nyc311 %>%
  select(Agency,
	 Complaint.Type,
	 Latitude,
   Longitude,
	 Borough) %>%
  filter(Agency=="HPD" & Borough!="Unspecified")
ggplot(data = locationData) +
      geom_point(mapping = aes(x = Longitude, y = Latitude,  color=Borough)) +
      ggtitle("Geo Plot for HPD Complaints")
```

# Conclusion
In this homework, I have gained a good understanding of the 311 NYC service call requests dataset. I have performed data pre-processing steps like removing irrelevant features for easier analysis and removing duplicates, included a data dictionary which I will be working on and explored the various relevant features of the service call requests data and depicted my findings by visualizing them with plots and tabulations. 


