---
title: 'homework iv'
author: 'Sachin Mohan Sujir'
date: '2020-09-25'
output: pdf_document
header-includes:
  \usepackage{booktabs}
---

```{r echo=FALSE}
# This chunk is just to make it possible to shrink the typeface in succeeding chunks. Mainly this will be used for the crosstabs.
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
```

# Introduction

In this document I am going to tidy the 311 data by removing errornous records in columnts like Incident Zip, Borought, Closed Date, etc. I will be determining the unspecified(missing) boroughs using the zip code provided for that record. Furthermore, I will be using the 'tidyr' functions like gather, spread, separate and complete to depict information in the form of tables which can be used for visualization purposes and to provide better insights. Finally, I will be introducing a relatable dataset that is relevant to the nyc311 dataset, which is NYPD NYC crimes data. I are taking a sample of around 95K from the original dataset which was around 5.5M (too large). Tidyr fuctions will be used to clean this dataset.

# Initialization

Here we load the tidyverse packages and the `data.table` package and load the nyc311 data set. Then we fix the column names of the nyc311 data so that they have no spaces.

```{r initialize}
library(tidyverse)
library(data.table)
nyc311<-fread("311_Service_Requests_from_2010_to_Present.csv", 
              na.strings = c("","NA","N/A"))
names(nyc311)<-names(nyc311) %>%
  stringr::str_replace_all("\\s", ".")
```

# Data pre-processing

Here we perform data pre-processing steps, by dropping irrelevant columns and removing duplicate rows from the nyc311 dataset.

```{r da taPreProcessing}
library(xtable)
options(xtable.comment=FALSE)
options(xtable.booktabs=TRUE)
nyc311<-nyc311 %>%
  select(Agency,
     Agency.Name,
     Created.Date,
     Closed.Date,
     Incident.Zip,
     Due.Date,
     Latitude,
     Longitude,
	 Complaint.Type,
	 Descriptor,

	 Status,
	 Borough)
xtable(head(nyc311))
nyc311 <- distinct(nyc311)
names(nyc311)
```

## Handling missing values

In the following snippet, I have handled the missing values and the errornous records in the columns of the data. 
Intially, I have replaced the invalid zip codes with NA if the zip code length is not 5 or 10 and if the zip code length is 10 then it should satisfy the "xxxxx-xxxx" format.
Besides, I could find zipcodes like 00000, 10000 which were invalid, hence replaced them with NA. 
Now considering the closed date column,there were dates that were defaulted to 01/01/1900 and also there were around 100K records with closed date lesser than the created date, which seems to be invalid and hence I replaced them with NA. 
For borough, there were around 800K records with unspecified values, out of which 600K had valid zip codes, so I found the boroughs for those records using the valid zipcode information and remaining was filled with NA. I could match the zip code that had missing borough and the zip code with the borough specified and filled the missing borough information. 

```{r handleMissingValues}
# Replacing invalid zipcodes with NA
nyc311[Incident.Zip=="00000" | (str_length(str_trim(Incident.Zip))<5 | 
        (str_length(str_trim(Incident.Zip)) > 5 & 
           str_length(str_trim(Incident.Zip)) < 10)  |                                                                                str_length(str_trim(Incident.Zip)) > 10 ) | 
          Incident.Zip=="10000","Incident.Zip"] <- NA

nyc311[as.Date(nyc311$Closed.Date, format="%m/%d/%Y")==
                as.Date("01/01/1900", format="%m/%d/%Y") | 
                as.Date(nyc311$Closed.Date, format="%m/%d/%Y")<
                  as.Date(nyc311$Created.Date, format="%m/%d/%Y"), 
             c("Closed.Date") ] <- NA

unspecifiedBro <- nyc311 %>%
  select(Incident.Zip, Borough) %>%
  filter(Borough=="Unspecified" & !is.na(Incident.Zip))

zipCodeTable <- nyc311 %>%
  select(Incident.Zip, Borough) %>%
  filter(Borough!="Unspecified" & (str_length(str_trim(Incident.Zip))==5 | 
   (str_length(str_trim(Incident.Zip))==10 & (str_detect(Incident.Zip,'-')))))
zipCodeTable <- distinct(zipCodeTable)
zipCodeTable <-  zipCodeTable %>%
 group_by(Incident.Zip) %>%
 summarize(Borough = first(Borough))

joinedTab <- merge(x=unspecifiedBro, y=zipCodeTable, by = "Incident.Zip", all.x = TRUE)
joinedTab <- distinct(joinedTab)
colnames(joinedTab)[colnames(joinedTab)=="Borough.x"] <- "Borough"

nyc311 <- merge(x=nyc311, y=joinedTab, 
                  by=c("Incident.Zip", "Borough"), sort=FALSE, all.x = TRUE)
nyc311[!is.na(Borough.y), "Borough"] <- nyc311[!is.na(Borough.y), "Borough.y"]
nyc311[Borough=="Unspecified", "Borough"] <- 
  nyc311[Borough=="Unspecified", "Borough.y"]
# drop the borough.y
nyc311 <- nyc311[,-"Borough.y"]
head(nyc311)
```

# Usage of TidyR

In the following snippet, I will be showing a table which depicts the frequency of complaints across every borough with respect to every complaint type. This was done by using spread function on the borough column.

```{r complaintBorough}
subsetData <- select(nyc311, Complaint.Type, Borough)
subsetData <- subsetData %>%
    filter(!is.na(Borough)) %>%
  group_by(Complaint.Type,Borough) %>%
  summarize(count=n())
newData <- complete(subsetData, Complaint.Type, Borough)
boroughSpread <- newData %>%
  spread(key=Borough, value=count)
boroughSpread[is.na(boroughSpread)] <- 0 
boroughSpread
```

In the following snippet, I will be showing a table which depicts the frequency of complaints for the top 5 agencies with respect to every complaint type. I did this by using group by function which is similar to gather function in tidyr library.

```{r complaintAgency}
AgencyCount <- select(nyc311,Complaint.Type , Agency)
(agencyData <- AgencyCount %>%
  filter(Agency== "HPD" | Agency == "DOT" | Agency=="NYPD" 
         | Agency == "DEP" | Agency=="DSNY") %>%
  group_by(Complaint.Type,Agency) %>%
  summarize(frequency = n()))
```

In the following snippet, I will be showing a table which depicts the year wise frequency of complaints with respect to every borough. I have achieved this by using separate function to extract the year from the created date, after which I have spreaded across the year, thus computing the frequency of complaints for each borough.

```{r BoroughYear}
boroughYear <-nyc311 %>%
  select( Borough , Created.Date, Complaint.Type) %>%
  filter(!is.na(Borough))
yearData <- separate(boroughYear, Created.Date, into=c("month", "day", "year"), 
                     convert = T)
boroughYear <- yearData %>%
  group_by(year, Borough) %>%
  summarize(frequency=n())
(yearSpread <- boroughYear %>%
  spread(key=year, value=frequency))
```

In the following snippet, I am showing a table which depicts the frequency of complaints across every borough with respect to the top 5 agencies. I have achieved this using spread function on the borough column.

```{r AgencyBorough}
AgencyBorough <- select(nyc311,Agency, Borough)
AgencyBorough <- AgencyBorough %>%
  filter((Agency== "HPD" | Agency == "DOT" | Agency=="NYPD" 
          | Agency == "DEP" | Agency=="DSNY") 
         & !is.na(Borough)) %>%
  group_by( Agency,Borough) %>%
  summarize(count= n())
(AgencyBorough <- AgencyBorough %>%
  spread(key = Borough, value = count))
```

# Relatable data set - NYPD NYC Crimes data

## Description

I have used the NYPD NYC crimes data which is a sample of size approx 95K records taken from the original data source.This dataset includes all valid felony, misdemeanor, and violation crimes reported to the New York City Police Department (NYPD). I found this dataset not only relevant to nyc311  but also interesting. 

## Initialization

Here I load the NYC Crimes data set from the link as provided below and I fill the empty cells with NA.

```{r readCrimesData}
nycCrimes <- 
  fread("https://raw.githubusercontent.com/jamesjynus/Shiny/master/data/crime.csv", 
                   na.strings = c("","NA"))
```

## Data pre-processing of NYC Crimes data

Here, I removed the irrelevant columns and duplicate records in the data, fixed the column name for borough and I am showing the head and data dictionary.

```{r crimeDataPreProcessing}
names(nycCrimes)
library(xtable)
options(xtable.comment=FALSE)
options(xtable.booktabs=TRUE)
nycCrimes<-nycCrimes %>%
  select(Date,
         Time,Code,Offense,Status,Type,
         Boro,Latitude,Longitude,Latitude,
         Population,Year_Month_New)
xtable(head(nycCrimes))
nycCrimes1 <- distinct(nycCrimes)
colnames(nycCrimes1)[colnames(nycCrimes1)=="Boro"] <- "Borough"
nycCrimes1 <-  nycCrimes1[str_trim(Offense)!="",]
names(nycCrimes1)
head(nycCrimes1)
```

## Data Dictionary

* Date - Date on which crime happened in the format yyyy-mm-dd.

* Time - Time at which crime occured in the format hh:mm:ss.

*	Code - Unique code for every offense.

*	Offense - The description of the crime type(sub-categories of the crime).

*	Status - The status of the crime report submitted(Allowed values: COMPLETED , ATTEMPTED).

*	Type - The type of crime(Allowed types: FELONY, MISDEMEANOR,VIOLATION).

*	Borough - town/district of the NYC provided by submitter(Values: BRONX, BROOKLYN, MANHATTAN, QUEENS, STATEN ISLAND).

*	Latitude - Geo-based latitude of the incident location(Type: degrees).

*	Longitude – Geo-based longitude of the incident location(Type: degrees).

*	Population - The population of the Borough on the date of the crime.

*	Year_Month_New - Year and Month of the crime date in the format yyyy-mm.

## Usage of TidyR

In the following snippet, I am showing a table which depicts the frequency of crimes across every borough with respect to every crime type. I have done this by using spread function on the borough column.

```{r crimesTypeBoro}
subsetData <- select(nycCrimes1, Type, Borough)
subsetData <- subsetData %>%
    filter(!is.na(Borough)) %>%
  group_by(Type,Borough) %>%
  summarize(count=n())
newData <- complete(subsetData,Type, Borough)
boroughSpread <- newData %>%
  spread(key=Borough, value=count)
boroughSpread[is.na(boroughSpread)] <- 0 
boroughSpread
```

In the following snippet, I am showing a table which depicts the year wise frequency of crimes for each borough. I have achieved this by using the separate function to extract the year from the created date, and then I have spread across the year, thus computing the frequency of crimes for each borough.

```{r crimesBoroYear}
boroYear <-nycCrimes1 %>%
  select( Borough , Year_Month_New,Type) %>%
  filter(!is.na(Borough))
yearData <- separate(boroYear, Year_Month_New, into=c("year", "month"), convert = T)
boroYear <- yearData %>%
  group_by(year,Borough) %>%
  summarize(frequency=n())
(yearSpread <- boroYear %>%
  spread(key=year, value=frequency))
```

# Conclusion

In this document, I have introduced a new dataset: NYPD NYC Crimes data relateable to our 311NYC data. I have performed data cleaning by dropping the irrelevant columns, removing the duplicates and replacing the missing values on both the datasets like filing in the borough using the zip code match, validating zip codes and removing noisy closed date records. I have also made use of the tidyR functions, showing relevant information with respect to complaints and crimes in the form of tables.
